Biogeography of crop progenitors and wild plant resources in Late Pleistocene–Early Holocene West Asia, 21–8.3 ka

### 2025-01-04 analysis/paper.qmd
Improve PDF formatting

---
title: Biogeography of crop progenitors and wild plant resources in West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of 56 plant species found regularly in association with
  early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using hindcast data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
  the Early Holocene compared to present conditions, indicating [...]. 
  The modelled ranges predict the observed occurrence of these species on 
  archaeological sites with [low|medium|high] accuracy. The regional ubiquity of 
  species in the archaeological record is [not] correlated with the predicted 
  size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
bibliography: references.bib
execute: 
  echo: false
---

* [x] Finish modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [ ] Results
    - [ ] Model assessment
    - [ ] Hindcasting
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-occ-count
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
- [ ] Clean up package code
- [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

@@

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
One relevant early example is @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites in Europe and West Asia to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworksyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik? --> [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].

@@

We consider X<!-- TODO --> distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic sites in West Asia, according to our previous study [@ArranzOtaeguiRoe2023].
<!-- TODO: something about not distinguishing between source of plant remains? -->

@@

Taxonomic names were resolves to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.

@@

```{r data-occ-count}
if (!file_exists(derived_data("flora_occ_count.tsv"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  write_tsv(flora, derived_data("flora_occ_count.tsv"))
}
flora <- read_tsv(derived_data("flora_occ_count.tsv"))
```

@@

```{r fetch-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  )
  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}
flora <- nest(read_tsv(derived_data("flora_occ.tsv")), occ = gbif_key:latitude)
```

@@

Predictor data was transformed to the same projection system (WGS84 / UTM 37 N) and a common resolution of X<!-- TODO --> km.

@@

For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our X<!--TODO--> predictor variables (@tbl-predictors).
We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (3/4) and test (1/4) partitions.

@@

```{r fig-model-accuracy-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
flora |>
  mutate(niche_test_fit = map(niche, collect_metrics)) |>
  unnest(niche_test_fit) |>
  filter(.metric == "accuracy") |> # TODO: or roc_auc?
  mutate(label = if_else(.estimate < 0.8, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
  theme_half_open() 
```

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in Late Pleistocene–Early Holocene West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of 56 plant species found regularly in association with
  early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using hindcast data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
  the Early Holocene compared to present conditions, indicating [...]. 
  The modelled ranges predict the observed occurrence of these species on 
  archaeological sites with [low|medium|high] accuracy. The regional ubiquity of 
  species in the archaeological record is [not] correlated with the predicted 
  size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Finish modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [ ] Results
    - [ ] Model assessment
    - [ ] Hindcasting
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
- [ ] Clean up package code
- [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

@@

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworksyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik? --> [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].

@@

@@

We began by considering `r nrow(flora)` distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic (c. 11.7–6.5 ka) sites in West Asia, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
<!-- TODO: something about not distinguishing between source of plant remains? -->
Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.

@@

```{r data-occ-count}
if (!file_exists(derived_data("flora_occ_count.tsv"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  write_tsv(flora, derived_data("flora_occ_count.tsv"))
}
flora <- read_tsv(
  derived_data("flora_occ_count.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_double()
  )
)
```

@@

```{r fetch-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  )
  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}
flora <- read_tsv(
  derived_data("flora_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_double(),
    gbif_key = col_double(),
    scientific_name = col_character(),
    genus = col_character(),
    species = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
) |>
  nest(occ = gbif_key:latitude)
```

@@

Predictor data was transformed to common equal-area projection and resolution of `r pred_res / 1000` km.

```{r tbl-climate-periods}
#| fig-cap: Palaeoclimatic periods used for hindcasting, after @BrownEtAl2018
# TODO: add refs (as footnotes?)
climate_periods |>
  mutate(
    code = toupper(code),
    age_start = start_bp / 1000,
    age_end = end_bp / 1000
  ) |>
  select(period, code, age_start, age_end) |>
  gt() |>
  cols_label(period = "Period", age_start = "Age") |>
  cols_merge(
    c(period, code),
    pattern = "{1} ({2})"
  ) |>
  cols_merge_range(age_start, age_end) |>
  cols_units(age_start = "ka") |>
  sub_missing()
```

@@

For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our X<!--TODO--> predictor variables (@tbl-predictors).
We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (¾) and test (¼) partitions.

@@

```{r fig-model-accuracy-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
flora |>
  mutate(niche_test_fit = map(niche, collect_metrics)) |>
  unnest(niche_test_fit) |>
  filter(.metric == "accuracy") |> # TODO: or roc_auc?
  mutate(label = if_else(.estimate < 0.75, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
  theme_half_open() 
```
